---
title: "Advent of Code Day 6"
author: "Dr Jeremiah MF Kelly"
date: "2024-12-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
```

## source
Navigation puzzle so lets sort the map

With matrix addresses 

```{r mats}
# Initialize variables
imax <- 10
jmax <- 10

maplet <- as.matrix(read.fwf("test.txt", widths = rep(1, imax), comment.char = ""))
loc <- which(maplet == "^", arr.ind = TRUE)  # Starting location
route <- character(0)
dir <- "N"
move <- list("N" = c(-1, 0), "S" = c(1, 0), "W" = c(0, -1), "E" = c(0, 1))

# Initialize path and other variables
path <- matrix(loc, ncol = 2, byrow = TRUE)
trip <- FALSE

# Direction rotation helper
rotate_dir <- function(current_dir) {
  directions <- c("N", "E", "S", "W")
  next_index <- (match(current_dir, directions) %% 4) + 1
  return(directions[next_index])
}

while (!trip) {
  # Calculate next step
  step_to <- loc + move[[dir]]
  
  # Check if out of bounds
  if (any(step_to <= 0) || any(step_to > imax)) {
    message("Elvish has left the building")
    trip <- TRUE
    break
  }
  
  # Check if path is blocked
  if (maplet[step_to[1], step_to[2]] == "#") {
    dir <- rotate_dir(dir)  # Change direction
  } else {
    # Move to the new location
    loc <- step_to
    path <- rbind(path, loc)
    route <- c(route, dir)
  }
}

# Convert coordinates to unique vector indices
vect <- path[, 1] + (path[, 2] - 1) * imax
unique_count <- length(unique(vect))

# Output results
cat("Unique positions visited:", unique_count, "\n")

```

## Part two 

Put an obstacle in so that the guard walks in circles

How many places can we put an obstacle to create a loop
What defines a loop 
if we put an obstacle at (7,4,)


The obstacle can only be on the original path, what we can do is put an object on the path and see is a loop follows
in a loop we get repeated patterns of location and direction

```{r loops}

# Initialize variables
orig_path <- path  # Original path
imax <- 130
jmax <- 130

reset <- as.matrix(read.fwf("input.txt", widths = rep(1, imax), comment.char = ""))
start_loc <- which(reset == "^", arr.ind = TRUE)  # Original starting location
loops <- 0

# Function to rotate direction
rotate_dir <- function(current_dir) {
  directions <- c("N", "E", "S", "W")
  next_index <- (match(current_dir, directions) %% 4) + 1
  return(directions[next_index])
}

# Loop through the original path
for (ii in seq_len(nrow(orig_path))) {
  # Skip if the obstacle placement is the original starting location
  if (all(orig_path[ii, ] == start_loc)) {
    next
  }

  # Reset starting position and map
  loc <- which(reset == "^", arr.ind = TRUE)
  maplet <- reset
  maplet[orig_path[ii, 1], orig_path[ii, 2]] <- "#"  # Place obstacle

  route <- character(0)
  dir <- "N"
  move <- list("N" = c(-1, 0), "S" = c(1, 0), "W" = c(0, -1), "E" = c(0, 1))
  path <- matrix(loc, ncol = 2, byrow = TRUE)
  trip <- FALSE
  steps <- 0
  places <- integer(0)

  # Simulate movement
  while (TRUE) {
    steps <- steps + 1
    step_to <- loc + move[[dir]]

    # Check if Elvish has left the building
    if (any(step_to <= 0) || any(step_to > imax)) {
      message("Elvish has left the building")
      trip <- TRUE
      break
    }

    # Check if the path is blocked
    if (maplet[step_to[1], step_to[2]] == "#") {
      dir <- rotate_dir(dir)  # Change direction
    } else {
      # Move to the new location
      loc <- step_to
      path <- rbind(path, loc)
      route <- c(route, dir)
      maplet[step_to[1], step_to[2]] <- "o"

      # Check for loop
      vect <- path[, 1] + (path[, 2] - 1) * imax
      unique_places <- length(unique(vect))
      places <- c(unique_places, places)

      # Detect loop
      if (steps > 1000 && length(places) > 1 && places[1] == places[2]) {
        loops <- loops + 1
        message("It's a loop!")
        break
      }
    }
  }
}

# Output the result
cat("Number of loops detected:", loops, "\n")



```


